---
title: "BiocOncoTK: a set of cancer-oriented software components for Bioconductor"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{BiocOncoTK -- cancer oriented components for Bioconductor}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::pdf_document:
    toc: yes
    number_sections: yes
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
suppressMessages({
library(BiocOncoTK)
library(BiocStyle)
})
})
```

# Overview

This package provides a unified approach to programming with Bioconductor
components to address problems in cancer genomics.  Central concerns are:

- __Ontology applications__ that systematize the conceptual structure of
cancer biology.  There are particular concerns with structured vocabularies for
    - general human anatomy 
    - experimental cell lines
    - cancer therapeutics
    - genome and epigenome elements and alterations related to cancer
- __Communications infrastructure__ to support extraction and analysis of well-structured, self-describing
data from public archives and portals.  Key resource centers of interest include
    - [TCGA](https://cancergenome.nih.gov/), with curated representations through
        - the [ISB Cancer Genomics Cloud project](http://cgc.systemsbiology.net/)
        - Bioconductor's [curatedTCGAData](https://bioconductor.org/packages/curatedTCGAData/) interface
    - [cBioPortal](http://www.cbioportal.org/); our main concern is to simplify usage of the RESTful API with R
    - [Ivy Glioblastoma Atlas Project](http://glioblastoma.alleninstitute.org/)
    - [TCIA](http://www.cancerimagingarchive.net/), with radiology and pathology
components available through [ISB/BigQuery](http://isb-cancer-genomics-cloud.readthedocs.io/en/latest/sections/TCGA-images.html)
    - [TCRN](https://cancerdatanetwork.org/): the TIES (Text Information Extraction System) Cancer Research Network -- we'll have to learn more about [JWT with httr](https://cran.r-project.org/web/packages/jose/vignettes/jwt.html) before progressing with this
    - [CONQUER](http://imlspenticton.uzh.ch:3838/conquer/), an archive of uniformly processed __single-cell RNA-seq__ datasets, a number of which related to cancer.  We have 'curated' CONQUER's version of the GBM
single cell study of [Patel et al. 2014](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4123637/)
in the [patelGBMSC package](https://github.com/vjcitn/patelGBMSC), with the large data component
in an AWS S3 bucket.
- __Remote analysis support__ to "move the computation to the data".  Bridging [Google Cloud Datalab](https://cloud.google.com/datalab/) to the ISB TCGA and CCLE images is a first target.
- __Social coding practice__: at present, users and developers should use the [issue tracker](https://github.com/vjcitn/BiocOncoTK/issues)
at the [BiocOncoTK github repository](https://github.com/vjcitn/BiocOncoTK) to comment, critique, and propose new approaches.

# Ontology

## Oncotree

The NCI Thesaurus project distributes an OBO representation of oncotree.  We
can use this through the `r Biocpkg("ontoProc")` (devel branch only) and `r CRANpkg("ontologyPlot")`
packages.  Code for visualizing the location of 'Glioblastoma' in the context of its 'siblings'
in the ontology follows.

```{r lkgbm,fig=TRUE,message=FALSE}
library(ontoProc)
library(ontologyPlot)
oto = getOncotreeOnto()
glioTag = names(grep("Glioblastoma$", oto$name, value=TRUE))
st = siblings_TAG(glioTag, oto, justSibs=FALSE)
onto_plot(oto, st@ontoTags, fontsize=50)
```


